Challenges to transboundary fisheries management in North America under climate change

### 2020-11-21 Ecol/Manuscript_Draft.Rmd
Update to manuscript

The shifts in distribution of shared fish stocks will impact the economics of their fisheries [@Pinsky:2012kq; @Sumaila:2019gh, @Sumaila:2020tp], and create international disputes between countries [@Munro:2002uf; @Spijkers:2017ij; @Pinsky:2018cb]. Canada and the US share important transboundary stocks of salmon (*Oncorhynchus spp.*), Pacific halibut (*Hippoglossus stenolepis*), and Atlantic cod (*Gadus morhua*) offering a unique lens to understand the extent to which climate-induced distributional shifts will challenge the future sustainability of transboundary fisheries. These countries have a long history of fisheries cooperation participating in diverse, jointly managed, commercial transboundary stocks through various fisheries management organizations [@NOAA:2018tu]. Furthermore, climate-related fluctuations in stocks' distribution have historically created disputes between Canada and the US, increasing international conflict and threatening the health of diverse transboundary stocks [@Munro:2002uf; @CIA:2017uta].

@@

The current study used the International Pacific Halibut Commission (IPHC) and the Gulf of Maine arrangement (hereafter referred as GoMA) as case studies to discuss the implications that climate change could have in the management of transboundary stocks. For the IPHC, we used the most updated spatial regulatory data along its 12 regulatory areas [@IPHCSecreatriat:2017uz; @IPHC:2019wt]. For this specific case, we considered Alaska as a separate entity, the US contiguous states as a second one (Washington, Oregon and California), and lastly British Columbia (Canada). For the GoMA we used the Northwest Atlantic Fisheries Organization's (NAFO) divisions 5Y, 5Ze, and 4X[^NAFO] within latitudes 46.2°N and 41.5°S, and longitudes -72°W and -64°E (Figure \ref{fig:Figure1-map}). Is worth mention that, while NAFO's divisional zones were used in this study for domestic management, NAFO does not manage fisheries within the EEZs of Canada and the United States. Fisheries data fpr the projections was gathered from the Sea Around Us from 1951 to 2014 [@Zeller2016].

@@

```{r Figure1-map, eval = T, echo=FALSE, fig.height = 4, fig.width = 6, fig.cap="Map of Canada and the US with the regulatory areas of the International Pacific Halibut Commission and the NAFO sub-divisions containing the Gulf of Maine arrangement"}

@@

Since 1998 Canada and the US have used a "Resource Sharing Understanding" to inform the management of Eastern George Bank's Atlantic cod (*Gadus morhua*) and haddock (*Melanogrammus aeglefinus*) and, yellowtail flounder (*Limanda ferruginea*) [@Pudden:2007bo; @TRAC:2016wu; @Song:2017va]. From 2010 onward, the GoMA suggests catch-limits based on a weighted method where 10% represents the stocks’ historical distribution (from 1967 to 1994) and 90% current distributions produced by quarterly surveys and catch [@TRAC:2016wu]. Since its introduction, the average quota allocation for each species proposed by the GoMA has been (Table \ref{tab:Table1}); Atlantic cod 77% Canada and 23% US, haddock 55% Canada and 45% US, and yellowtail flounder 34% Canada and 66% US [@TRAC:2015wj; @TRAC:2015vp; @TRAC:2015tt; @TRAC:2018tw; @TRAC:2018tj; @TRAC:2018wx; @Trinko:2019ou]. However, because this is an unofficial agreement, Canada and the US ultimately take single management decisions [@Soboil:2006gw]. In terms of management, the US has a multi species harvest control with area and season closures, mesh sizes, effort control, and mobile gear vessels that use bottom otter trawl gear [@Soboil:2006gw]. In contrast, Canada has a quota system in addition to limited-entry licensing, fleet allocations, and mesh and fish size regulation, among other input controls. Canada inshore vessels fish cod with longline and gillnet while haddock is mainly caught with bottom otter trawl gear [@Soboil:2006gw].

```{r Table1, eval = T, echo = F, warning=F, message=F}

@@

where *y* is year, *r* is region, *s* is species, *n* is total number of species, and $\hat{MCP}$ is the MCP averaged by the three ESMs. In the case of the GoMA, region was defined as the $0.5^{\circ}$ x $0.5^{\circ}$ grid-cell within the specific NAFO divisional zones. For the IPHC analysis, region was defined as the Commission's regulatory areas (Figure \ref{fig:Figure1-map}). We then averaged the values in three time periods (*t*) to reduce temporal model sensitivity. Thus, we computed the regional percentage change in MCP ($\Delta MCP$) as follows:

@@

Where $\theta_{rt}$* is the species' aggregated $\hat{MCP}$ of each region at time period *t*, and $\delta_{rt}$ is the species' aggregated $\hat{MCP}$ of the whole specie's distribution within the study area at the same time period. Finally, we estimated the percentage change in stock-share ratio substituting $X_{t_0}$ and $X_{t}$ by $\alpha_{t0}$ and $\alpha_t$, respectively in Equation \@ref(eq:two). The process was carried out for each ESM and results presented as average $\pm$ standard deviation (sd). All of the analysis was done in the statistical software R version 3.5.2 (2018-12-20) with the associated packages, *data.table* [@Packagedatatable:2019uh], *ggrepel* [@PackageggrepelAut:2018to], *gridExtra* [@PackagegridExtraM:2017wx], *knirt* [@PackageknitrAGen:2020wt], *RColorBrewer* [@PackagerColorBrewer:2014uw], *sf* [@PackagesfSimpleF:2018vp], and *tidyverse* [@PackagetidyverseE:2017vq]. All code is available at http://www.github.com/jepa/OC_Transboundary.

@@

At least one third of the IPHC regulatory areas will see a reduction in MCP of Pacific halibut by 2050 relative to current MCP, regardless of the climate change scenario (Figure \ref{fig:Figure2-IPHC}). It is likely that the stock shift from the US. contiguous states towards Canada will offset the shift of the later towards northern regions, resulting in undetectable changes in Canadian area 2B and Alaskan 2C under both climate change scenarios. The potential movement of halibut westward will increase the MCP of regulatory areas 3B (under a low emission scenario) and 4ABCE along the Aleutian Islands and Bering Sea. Regions 4DE, the most poleward regulatory areas of the IPHC, are expected to gain MCP by mid (Figure \ref{fig:Figure2-IPHC}) and end of the $21^{st}$ century (Figure \ref{fig:FigureS11}) under a high emission scenario due to the expansion of halibut suitable habitat as sea ice retreats (Figure \ref{fig:FigureS21}). In contrast, under a low emission scenario, sea ice is expected to stabilize towards mid 21st century, thus providing less “new” suitable habitat for Pacific halibut and resulting in undetectable changes in MCP for the region (Figure \ref{fig:Figure2-IPHC}B) and decreasing even more towards 2100 (Figure \ref{fig:FigureS11}).

@@

IPHC_DBEM_sf <- IPHC_DBEM_sf %>% 
 # filter(Period == "Mid_Century") %>% # Paper figure
 filter(Period == "The_End_Century") %>%# Supplemental figure
 mutate(RegArea = paste(RegArea,Model_Agreement,sep=""))

@@

```{r Figure2-IPHC, eval = T, echo = F, fig.cap= "Percentage change of MCP for species managed by the IPHC for mid 21st century (2041-2060) relative to present 2005-2014 under a A) high emission scenario and B) low emission scenario. Labels marked with * represent regions where models do not agree in direction of change."}

@@

The same poleward trend is expected in the change of Pacific halibut stock-share ratio with the average proportion increasing up to 25% in some northern regions and decreasing by 10% in southern regions, relative to the present proportion (Figure \ref{fig:Figure3-IPHC-prop}). Maintaining emissions to lower levels through 2050 would potentially leave unchanged the stock-share ratio of three regulatory areas (3AC, and 4D) and negatively change regulatory area 2A. On the other hand, failing to achieve such target will decrease the stock-share ratio in the most productive regulatory areas (2AC, 3AB).

@@

```{r Figure3-IPHC-prop, eval = T, echo = F, fig.cap="Percentage change of stock-share ratio for IPHC under A) high emission scenario and B) low emission scenario for mid 21st century (2041-2060) relative to present 2005-2014. Values represent the mean of 3 ESM; error bars represent \u00B1 sd."}

@@

While some regulatory areas of the IPHC will see an incremental increase in Pacific halibut MCP, the results for the Gulf of Maine show an overall decrease in MCP by 2050, regardless of the climate change scenario or ESM (Figure \ref{fig:Figure4-GOMA}), intensifying by the end of the $21^{st}$ century (Figure \ref{fig:FigureS12}). For cod and haddock, MCP will decrease within the whole Gulf with no apparent win for any country in reference to the current period (Figure \ref{fig:Figure4-GOMA}). For yellowtail flounder, despite an overall reduction, some discrete areas are expected to increase with no particular pattern and high uncertainty, as ESMs in these regions do not agree in the direction of change (Figure \ref{fig:FigureS23}). Despite the overall reduction in MCP for all three species in comparison to current values, there is a benefit of achieving a low emission scenario as reductions intensify under the high emission scenario.

@@

```{r Figure4-GOMA, eval = T, echo = F, fig.cap= "Percentage change of MCP in the Gulf of Maine under, A) high emission scenario; and B) low emission for mid-century (2041-2060) relative to present (2005-2014). Points represent regions where models do not agree in direction of change."}

@@

Despite the expected decrease in MCP for the region, changes in the stock-share ratio of species within the Gulf of Maine show different outcomes dependent on the climate change scenario and species in question. Following a high emission path will affect mostly Canada’s share of yellowtail flounder and in less degree haddock, with an increase of cod share. under the low emission scenario, haddock and cod patterns intensify, while yellowtail flounder’s share approaches almost no change (Figure \ref{fig:Figure5-GOMA-prop}). Such pattern is likely the combination of the bathymetry or the Gulf, the warming gradient, and the species distribution (see Discussion).

@@

```{r Figure5-GOMA-prop, eval = T, echo = F, fig.cap="Changes in MCP stock-share ratio for Gulf of Maine under (A) high emission; and (B) low emission scenarios for mid 21st century (2041-2060) relative to present (2005-2014). Values represent the mean of 3 ESM; error bars represent \u00B1 sd."}

@@

The results of the present study suggest that climate change will alter the MCP of jointly managed transboundary fish stocks in North America consequently altering Canada’s and the US’s species’ stock-share ratio, regardless of the climate change scenario. These results are aligned with regional [@Morley:2018fn] projections suggesting that climate change will push marine species towards the poles and deeper water [@Pinsky:2013ab] in search of their ecological niche [@Poloczanska:2016kk]. Moreover, IPHC data[^IPHC] suggest that some of these shits are already happening. For example, since 2010, the distribution proportion of Pacific halibut has increased from 9% to 11% in region 2B, from 7.5% to 13% in region 2C, and from 12.3% to 13.5% in region 4CDE. On the other hand, regions 3A and 3B have seen the largest decreases in the IPHC regulatory areas since 2010, from 35.3% to 30.6% and 20.6% to 15.9%, respectively. Similarly, in the Gulf of Maine, the projected stock-share gain of yellowtail flounder and haddock by the US (Figure \ref{fig:Figure5-GOMA-prop}) follows a historical trend where in 2019, Canada's stock-share decreased from 35% to 32% and 60% to 40% relative to 2010, respectively [@Trinko:2019ou].

@@

Geographic barriers [@Cheung:2015fm; @Kleisner:2016je], local temperature gradients [@Pinsky:2013ab], species interactions and human activities [@Serpetti:2017he] might change the rate and direction of species shifts. For the IPHC, geographic barriers might induce a westward increase of stock-share in IPHC regions where species can only migrate northward into the Arctic Ocean through the Bering Sea and Bering Strait [@Cheung:2015fm] (Figure \ref{fig:Figure3-IPHC-prop}). In the Gulf of Maine, future projections could be a response to a temperature gradient shift combined with geographic barriers as southern waters are deeper and warming slower than northern waters according to the ESMs (Figure \ref{fig:FigureS22}). Moreover, Maine has seen its landings of yellowtail flounder increased at the expenses of southern states [@Pinsky:2012kq]. This could be influencing the US gain in MCP in the GoMA in relation to Canada as species shift their distribution from lower latitudes naturally reaching the US (lower) region first. As the effects of climate change endure, even with high mitigation, joint plans should prepare to face changes in the stock-share ratio of transboundary stocks along both coast of North America.

The shifts in the distribution of transboundary stocks can jeopardize management objectives such as conservation measures and gear operation. Fish moving out of fishing grounds and into protected areas could result in a pressure increase to open such area to fishing. Moreover, overlapping shifting stocks could interfere in gear-limitation management rules of multiple fisheries generating conflicts between fleets [@VanDerVoo:2016uc]. The effectiveness of the IPHC-Closed Area ("CA" in Figure \ref{fig:Figure2-IPHC}) in terms of protecting juveniles has been historically questioned as trawling for other species is still allowed in the area [@Karim:2010ut; @IPHC:2017CA]. In 2015, for example, 97% of the trawl by-catch in areas 4CDE and the Closed Area were juveniles [@IPHC:2017CA]. Consequently, the Alaskan trawl fisheries has been closed before reaching annual quota due to the attainment of Pacific halibut bycatch quota limits [@Karim:2010ut]. Thus, the commission has been asked to open the closed area for Pacific halibut fishing, under the premise that the expansion of the trawl fishery is likely reducing any conservation goal for juvenile Pacific halibut [@IPHC:2017CA]. Although not assessed in this study, some trawling target species like Pacific cod (*Gadus macrocephalus*), flathead sole (*Hippoglossoides elassodon*), and Alaskan plaice (*Pleuronectes quadrituberculatus*) have already shift their distributions due to warming waters [@Stram:2009xc] and some are expected to continue shifting in similar direction than Pacific halibut [@Pinsky:2013ab]. The overlap of target species could be addressed by applying ecosystem-based management and dynamic management tools [@Hazen:2018fa] to manage these fisheries and reduce potential loss of sustainable harvest for both the halibut and the trawl fisheries.

@@

Transboundary fisheries management have to be prepared for the uncertainty that comes with a changing world. Future climate change will depend on the path society as a whole will take, and thus we rely on scenario planning to account for the uncertainty build in future decision making [@vanVuuren:2011ef]. In our results, the "winners and losers" of climate change, and the intensity of the change, will be scenario dependent. For instance, stock-share of yellowtail flounder under a high emission scenario will be larger for the US while Canada's gain of cod stock-share will be larger under a low emission scenario (Figure \ref{fig:Figure5-GOMA-prop}). Applying previously describe strategies (e.g. quota swaps or EEZ-fishing rights) could increase the resilience of treaties by preventing members from leaving the agreement due to a shift in threat points, as happened in the Pacific salmon case [@Miller:2013iv].

Models are attempts to represent reality (in our case a future reality) based on observational data, previously established theory, and future scenarios, and are thus, subjected to different degrees of uncertainty [@Payne:2016gn]. An ensemble of models is a way to present a more robust result that accounts for differences in the structural composition of each model [@Cheung:2016gv]. In here we used three ESMs to project future changes in species maximum catch potential. The levels of uncertainty related to the ESMs differ among case studies. Overall, results for the Gulf of Maine agree with a reduction in MCP of all three species. However, some discrete areas show a positive change for yellowtail flounder by mid $21^{st}$ century (Figure \@ref(fig:Figure3-IPHC-prop)), mainly driven by the GFDL model (Figure \@ref(fig:FigureS23)). Potential model artifacts could also be contributing to the results, especially in the northern part of the study area (Bay of Fundy) as most disagreeing grids are covered by land, which could be influencing the results. In contrast, considerable uncertainty exists in the change of MCP along the IPHC Convention area shown by a disagreement between ESMs (Figure \@ref(fig:FigureS24)). Off the coast of British Columbia, increasing temperature trends are consistent among ESMs, however, other processes such as acidification and deoxygenation are still not well understood from British Columbia to the Gulf of Alaska  [@TalloniAlvarez:2019gz]. Moreover, considerable uncertainty exists along the Bering Sea [@Douglas:2010Mb] and Antarctic Pacific regarding the extent and intensity of future sea-ice reduction under climate change [@Steiner:2015hc; @IPCC:2019tn]. Regarding the DBEM, its structural uncertainty has been previously tested for agreement against commonly used species distribution algorithms such as Maxent [@Phillips:2006ff] and AquaMaps [@Kaschner:2011hi; @Ready:2010wr] resulting in no qualitative differences in trends between algorithms [@Cheung:2016jd]. Finally, is worth mentioning that future changes to species distributions could be influenced by factors not captured by our model such as interactions between species [@Pecl:2017hu], adaptation of species to environmental changes, and anthropogenic factors [@Serpetti:2017he]. However, these factors are expected to increase the rate of range-shifting of the species making our results conservative [@Cheung:2010dt; @Serpetti:2017he].

@@

## Supplements

@@

```{r FigureS11, eval = T, echo = F, fig.cap = "Percentage change of MCP for species managed by the IPHC for end-century (2081–2100) relative to 2005–2014 under a A) high emission scenario and B) low emission scenario. Labels marked with * represent regions where models do not agree in direction of change."}

@@

```{r FigureS12, eval = T, echo = F, fig.cap = "Percentage change of MCP in the Gulf of Maine under (RCP 8.5) high emission scenario and (RCP 2.6) low emission for end-century (2080–2100) referent to present (2005– 2014). Values represent the mean of 3 ESM. Points represent regions where ESMs do not agree in direction of change"}

@@

```{r FigureS21, eval = T, echo = F, fig.cap = "Projected environmental variables under climate change from 2010 to 2010 for Arctic regions of the IPHC (Top:4D, Bottom, 4E). The solid line represents the average of all three ESMs and the shaded area represents the model’s uncertainty (s.d.)."}

@@

```{r FigureS22, eval = T, echo = F, fig.cap = "Depth profile (A) and bottom water warming of the Gulf of Maine. Everything deeper than 1000 meters is colored in green. B) Percentage change of bottom temperature relative to the present showing more intense warming in northern regions, especially under a high emission climate change scenario."}

@@

```{r FigureS24, eval = T, echo = F, fig.cap = "Projected MCP change relative to present (2005-2014) for each IPHC regulatory area. Colors represents the different ESM used in the study. Solid line represents a high emission scenario (RCP 8.5) and dashed line represents a low emission scenario (RCP 2.6)"}

@@

```{r Figure_out_outliers, eval = F, echo = F}

>>>>

The shifts in distribution of shared fish stocks will impact the economics of their fisheries [@Pinsky:2012kq; @Sumaila:2019gh; @Sumaila:2020tp], and create international disputes between countries [@Munro:2002uf; @Spijkers:2017ij; @Pinsky:2018cb]. Canada and the US share important transboundary stocks of salmon (*Oncorhynchus spp.*), Pacific halibut (*Hippoglossus stenolepis*), and Atlantic cod (*Gadus morhua*) offering a unique lens to understand the extent to which climate-induced distributional shifts will challenge the future sustainability of transboundary fisheries. These countries have a long history of fisheries cooperation participating in diverse, jointly managed, commercial transboundary stocks through various fisheries management organizations [@NOAA:2018tu]. Furthermore, climate-related fluctuations in stocks' distribution have historically created disputes between Canada and the US, increasing international conflict and threatening the health of diverse transboundary stocks [@Munro:2002uf; @CIA:2017uta].

@@

The current study used the International Pacific Halibut Commission (IPHC) and the Gulf of Maine arrangement (hereafter referred as GoMA) as case studies to discuss the implications that climate change could have in the management of transboundary stocks. For the IPHC, we used the most updated spatial regulatory data along its 12 regulatory areas [@IPHCSecreatriat:2017uz; @IPHC:2019wt]. For this specific case, we considered Alaska as a separate entity, the US contiguous states as a second one (Washington, Oregon and California), and lastly British Columbia (Canada). For the GoMA we used the Northwest Atlantic Fisheries Organization's (NAFO) divisions 5Y, 5Ze, and 4X[^NAFO] within latitudes 46.2°N and 41.5°S, and longitudes -72°W and -64°E (Figure \ref(fig:ocFigure1-map)). Is worth mention that, while NAFO's divisional zones were used in this study for domestic management, NAFO does not manage fisheries within the EEZs of Canada and the United States. Fisheries data fpr the projections was gathered from the Sea Around Us from 1951 to 2014 [@Zeller2016].

@@

```{r ocFigure1-map, eval = T, echo=FALSE, fig.height = 4, fig.width = 6, fig.cap="Map of Canada and the US with the regulatory areas of the International Pacific Halibut Commission and the NAFO sub-divisions containing the Gulf of Maine arrangement"}

@@

Since 1998 Canada and the US have used a "Resource Sharing Understanding" to inform the management of Eastern George Bank's Atlantic cod (*Gadus morhua*) and haddock (*Melanogrammus aeglefinus*) and, yellowtail flounder (*Limanda ferruginea*) [@Pudden:2007bo; @TRAC:2016wu; @Song:2017va]. From 2010 onward, the GoMA suggests catch-limits based on a weighted method where 10% represents the stocks’ historical distribution (from 1967 to 1994) and 90% current distributions produced by quarterly surveys and catch [@TRAC:2016wu]. Since its introduction, the average quota allocation for each species proposed by the GoMA has been (Table \@ref(tab:ocTable1); Atlantic cod 77% Canada and 23% US, haddock 55% Canada and 45% US, and yellowtail flounder 34% Canada and 66% US [@TRAC:2015wj; @TRAC:2015vp; @TRAC:2015tt; @TRAC:2018tw; @TRAC:2018tj; @TRAC:2018wx; @Trinko:2019ou]. However, because this is an unofficial agreement, Canada and the US ultimately take single management decisions [@Soboil:2006gw]. In terms of management, the US has a multi species harvest control with area and season closures, mesh sizes, effort control, and mobile gear vessels that use bottom otter trawl gear [@Soboil:2006gw]. In contrast, Canada has a quota system in addition to limited-entry licensing, fleet allocations, and mesh and fish size regulation, among other input controls. Canada inshore vessels fish cod with longline and gillnet while haddock is mainly caught with bottom otter trawl gear [@Soboil:2006gw].

```{r ocTable1, eval = T, echo = F, warning=F, message=F}

@@

where *y* is year, *r* is region, *s* is species, *n* is total number of species, and $\hat{MCP}$ is the MCP averaged by the three ESMs. In the case of the GoMA, region was defined as the $0.5^{\circ}$ x $0.5^{\circ}$ grid-cell within the specific NAFO divisional zones. For the IPHC analysis, region was defined as the Commission's regulatory areas (Figure \ref(fig:ocFigure1-map)). We then averaged the values in three time periods (*t*) to reduce temporal model sensitivity. Thus, we computed the regional percentage change in MCP ($\Delta MCP$) as follows:

@@

Where $\theta_{rt}$* is the species' aggregated $\hat{MCP}$ of each region at time period *t*, and $\delta_{rt}$ is the species' aggregated $\hat{MCP}$ of the whole specie's distribution within the study area at the same time period. Finally, we estimated the percentage change in stock-share ratio substituting $X_{t_0}$ and $X_{t}$ by $\alpha_{t0}$ and $\alpha_t$, respectively in Equation \@ref(eq:two). The process was carried out for each ESM and results presented as average $\pm$ standard deviation (sd). All of the analysis was done in the statistical software R version 3.5.2 (2018-12-20) with the associated packages, data.table [@Packagedatatable:2019uh], ggrepel [@PackageggrepelAut:2018to], gridExtra [@PackagegridExtraM:2017wx], knirt [@PackageknitrAGen:2020wt], RColorBrewer [@PackagerColorBrewer:2014uw], sf [@PackagesfSimpleF:2018vp], and tidyverse [@PackagetidyverseE:2017vq]. All code is available at http://www.github.com/jepa/OC_Transboundary.

@@

At least one third of the IPHC regulatory areas will see a reduction in MCP of Pacific halibut by 2050 relative to current MCP, regardless of the climate change scenario (Figure \ref(fig:ocFigure2-IPHC)). It is likely that the stock shift from the US. contiguous states towards Canada will offset the shift of the later towards northern regions, resulting in undetectable changes in Canadian area 2B and Alaskan 2C under both climate change scenarios. The potential movement of halibut westward will increase the MCP of regulatory areas 3B (under a low emission scenario) and 4ABCE along the Aleutian Islands and Bering Sea. Regions 4DE, the most poleward regulatory areas of the IPHC, are expected to gain MCP by mid (Figure \ref(fig:ocFigure2-IPHC)) and end of the $21^{st}$ century (Figure \ref(fig:ocFigureS11)) under a high emission scenario due to the expansion of halibut suitable habitat as sea ice retreats (Figure \ref(fig:ocFigureS21)). In contrast, under a low emission scenario, sea ice is expected to stabilize towards mid 21st century, thus providing less “new” suitable habitat for Pacific halibut and resulting in undetectable changes in MCP for the region (Figure \ref(fig:ocFigure2-IPHC)B) and decreasing even more towards 2100 (Figure \ref(fig:ocFigureS11)).

@@

IPHC_DBEM_sf <- IPHC_DBEM_sf %>% 
 # filter(Period == "Mid_Century") %>% # Paper ocFigure
 filter(Period == "The_End_Century") %>%# Supplemental figure
 mutate(RegArea = paste(RegArea,Model_Agreement,sep=""))

@@

```{r ocFigure2-IPHC, eval = T, echo = F, fig.cap= "Percentage change of MCP for species managed by the IPHC for mid 21st century (2041-2060) relative to present 2005-2014 under a A) high emission scenario and B) low emission scenario. Labels marked with * represent regions where models do not agree in direction of change."}

@@

The same poleward trend is expected in the change of Pacific halibut stock-share ratio with the average proportion increasing up to 25% in some northern regions and decreasing by 10% in southern regions, relative to the present proportion (Figure \ref(fig:ocFigure3-IPHC-prop)). Maintaining emissions to lower levels through 2050 would potentially leave unchanged the stock-share ratio of three regulatory areas (3AC, and 4D) and negatively change regulatory area 2A. On the other hand, failing to achieve such target will decrease the stock-share ratio in the most productive regulatory areas (2AC, 3AB).

@@

```{r ocFigure3-IPHC-prop, eval = T, echo = F, fig.cap="Percentage change of stock-share ratio for IPHC under A) high emission scenario and B) low emission scenario for mid 21st century (2041-2060) relative to present 2005-2014. Values represent the mean of 3 ESM; error bars represent \u00B1 sd."}

@@

While some regulatory areas of the IPHC will see an incremental increase in Pacific halibut MCP, the results for the Gulf of Maine show an overall decrease in MCP by 2050, regardless of the climate change scenario or ESM (Figure \ref(fig:ocFigure4-GOMA)), intensifying by the end of the $21^{st}$ century (Figure \ref(fig:ocFigureS12)). For cod and haddock, MCP will decrease within the whole Gulf with no apparent win for any country in reference to the current period (Figure \ref(fig:ocFigure4-GOMA)). For yellowtail flounder, despite an overall reduction, some discrete areas are expected to increase with no particular pattern and high uncertainty, as ESMs in these regions do not agree in the direction of change (Figure \ref(fig:ocFigureS23)). Despite the overall reduction in MCP for all three species in comparison to current values, there is a benefit of achieving a low emission scenario as reductions intensify under the high emission scenario.

@@

```{r ocFigure4-GOMA, eval = T, echo = F, fig.cap= "Percentage change of MCP in the Gulf of Maine under, A) high emission scenario; and B) low emission for mid-century (2041-2060) relative to present (2005-2014). Points represent regions where models do not agree in direction of change."}

@@

Despite the expected decrease in MCP for the region, changes in the stock-share ratio of species within the Gulf of Maine show different outcomes dependent on the climate change scenario and species in question. Following a high emission path will affect mostly Canada’s share of yellowtail flounder and in less degree haddock, with an increase of cod share. under the low emission scenario, haddock and cod patterns intensify, while yellowtail flounder’s share approaches almost no change (Figure \ref(fig:ocFigure5-GOMA-prop)). Such pattern is likely the combination of the bathymetry or the Gulf, the warming gradient, and the species distribution (see Discussion).

@@

```{r ocFigure5-GOMA-prop, eval = T, echo = F, fig.cap="Changes in MCP stock-share ratio for Gulf of Maine under (A) high emission; and (B) low emission scenarios for mid 21st century (2041-2060) relative to present (2005-2014). Values represent the mean of 3 ESM; error bars represent \u00B1 sd."}

@@

The results of the present study suggest that climate change will alter the MCP of jointly managed transboundary fish stocks in North America consequently altering Canada’s and the US’s species’ stock-share ratio, regardless of the climate change scenario. These results are aligned with regional [@Morley:2018fn] projections suggesting that climate change will push marine species towards the poles and deeper water [@Pinsky:2013ab] in search of their ecological niche [@Poloczanska:2016kk]. Moreover, IPHC data[^IPHC] suggest that some of these shits are already happening. For example, since 2010, the distribution proportion of Pacific halibut has increased from 9% to 11% in region 2B, from 7.5% to 13% in region 2C, and from 12.3% to 13.5% in region 4CDE. On the other hand, regions 3A and 3B have seen the largest decreases in the IPHC regulatory areas since 2010, from 35.3% to 30.6% and 20.6% to 15.9%, respectively. Similarly, in the Gulf of Maine, the projected stock-share gain of yellowtail flounder and haddock by the US (Figure \ref(fig:ocFigure5-GOMA-prop)) follows a historical trend where in 2019, Canada's stock-share decreased from 35% to 32% and 60% to 40% relative to 2010, respectively [@Trinko:2019ou].

@@

Geographic barriers [@Cheung:2015fm; @Kleisner:2016je], local temperature gradients [@Pinsky:2013ab], species interactions and human activities [@Serpetti:2017he] might change the rate and direction of species shifts. For the IPHC, geographic barriers might induce a westward increase of stock-share in IPHC regions where species can only migrate northward into the Arctic Ocean through the Bering Sea and Bering Strait [@Cheung:2015fm] (Figure \ref(fig:ocFigure3-IPHC-prop)). In the Gulf of Maine, future projections could be a response to a temperature gradient shift combined with geographic barriers as southern waters are deeper and warming slower than northern waters according to the ESMs (Figure \ref(fig:ocFigureS22)). Moreover, Maine has seen its landings of yellowtail flounder increased at the expenses of southern states [@Pinsky:2012kq]. This could be influencing the US gain in MCP in the GoMA in relation to Canada as species shift their distribution from lower latitudes naturally reaching the US (lower) region first. As the effects of climate change endure, even with high mitigation, joint plans should prepare to face changes in the stock-share ratio of transboundary stocks along both coast of North America.

The shifts in the distribution of transboundary stocks can jeopardize management objectives such as conservation measures and gear operation. Fish moving out of fishing grounds and into protected areas could result in a pressure increase to open such area to fishing. Moreover, overlapping shifting stocks could interfere in gear-limitation management rules of multiple fisheries generating conflicts between fleets [@VanDerVoo:2016uc]. The effectiveness of the IPHC-Closed Area ("CA" in Figure \ref(fig:ocFigure2-IPHC)) in terms of protecting juveniles has been historically questioned as trawling for other species is still allowed in the area [@Karim:2010ut; @IPHC:2017CA]. In 2015, for example, 97% of the trawl by-catch in areas 4CDE and the Closed Area were juveniles [@IPHC:2017CA]. Consequently, the Alaskan trawl fisheries has been closed before reaching annual quota due to the attainment of Pacific halibut bycatch quota limits [@Karim:2010ut]. Thus, the commission has been asked to open the closed area for Pacific halibut fishing, under the premise that the expansion of the trawl fishery is likely reducing any conservation goal for juvenile Pacific halibut [@IPHC:2017CA]. Although not assessed in this study, some trawling target species like Pacific cod (*Gadus macrocephalus*), flathead sole (*Hippoglossoides elassodon*), and Alaskan plaice (*Pleuronectes quadrituberculatus*) have already shift their distributions due to warming waters [@Stram:2009xc] and some are expected to continue shifting in similar direction than Pacific halibut [@Pinsky:2013ab]. The overlap of target species could be addressed by applying ecosystem-based management and dynamic management tools [@Hazen:2018fa] to manage these fisheries and reduce potential loss of sustainable harvest for both the halibut and the trawl fisheries.

@@

Transboundary fisheries management have to be prepared for the uncertainty that comes with a changing world. Future climate change will depend on the path society as a whole will take, and thus we rely on scenario planning to account for the uncertainty build in future decision making [@vanVuuren:2011ef]. In our results, the "winners and losers" of climate change, and the intensity of the change, will be scenario dependent. For instance, stock-share of yellowtail flounder under a high emission scenario will be larger for the US while Canada's gain of cod stock-share will be larger under a low emission scenario (Figure \ref(fig:ocFigure5-GOMA-prop)). Applying previously describe strategies (e.g. quota swaps or EEZ-fishing rights) could increase the resilience of treaties by preventing members from leaving the agreement due to a shift in threat points, as happened in the Pacific salmon case [@Miller:2013iv].

Models are attempts to represent reality (in our case a future reality) based on observational data, previously established theory, and future scenarios, and are thus, subjected to different degrees of uncertainty [@Payne:2016gn]. An ensemble of models is a way to present a more robust result that accounts for differences in the structural composition of each model [@Cheung:2016gv]. In here we used three ESMs to project future changes in species maximum catch potential. The levels of uncertainty related to the ESMs differ among case studies. Overall, results for the Gulf of Maine agree with a reduction in MCP of all three species. However, some discrete areas show a positive change for yellowtail flounder by mid $21^{st}$ century (Figure \@ref(fig:ocFigure3-IPHC-prop)), mainly driven by the GFDL model (Figure \@ref(fig:ocFigureS23)). Potential model artifacts could also be contributing to the results, especially in the northern part of the study area (Bay of Fundy) as most disagreeing grids are covered by land, which could be influencing the results. In contrast, considerable uncertainty exists in the change of MCP along the IPHC Convention area shown by a disagreement between ESMs (Figure \@ref(fig:ocFigureS24)). Off the coast of British Columbia, increasing temperature trends are consistent among ESMs, however, other processes such as acidification and deoxygenation are still not well understood from British Columbia to the Gulf of Alaska  [@TalloniAlvarez:2019gz]. Moreover, considerable uncertainty exists along the Bering Sea [@Douglas:2010Mb] and Antarctic Pacific regarding the extent and intensity of future sea-ice reduction under climate change [@Steiner:2015hc; @IPCC:2019tn]. Regarding the DBEM, its structural uncertainty has been previously tested for agreement against commonly used species distribution algorithms such as Maxent [@Phillips:2006ff] and AquaMaps [@Kaschner:2011hi; @Ready:2010wr] resulting in no qualitative differences in trends between algorithms [@Cheung:2016jd]. Finally, is worth mentioning that future changes to species distributions could be influenced by factors not captured by our model such as interactions between species [@Pecl:2017hu], adaptation of species to environmental changes, and anthropogenic factors [@Serpetti:2017he]. However, these factors are expected to increase the rate of range-shifting of the species making our results conservative [@Cheung:2010dt; @Serpetti:2017he].

@@

<!-- ## Supplements -->

@@

```{r ocFigureS11, eval = F, echo = F, fig.cap = "Percentage change of MCP for species managed by the IPHC for end-century (2081–2100) relative to 2005–2014 under a A) high emission scenario and B) low emission scenario. Labels marked with * represent regions where models do not agree in direction of change."}

@@

```{r ocFigureS12, eval = F, echo = F, fig.cap = "Percentage change of MCP in the Gulf of Maine under (RCP 8.5) high emission scenario and (RCP 2.6) low emission for end-century (2080–2100) referent to present (2005– 2014). Values represent the mean of 3 ESM. Points represent regions where ESMs do not agree in direction of change"}

@@

```{r ocFigureS21, eval = F, echo = F, fig.cap = "Projected environmental variables under climate change from 2010 to 2010 for Arctic regions of the IPHC (Top:4D, Bottom, 4E). The solid line represents the average of all three ESMs and the shaded area represents the model’s uncertainty (s.d.)."}

@@

```{r ocFigureS22, eval = F, echo = F, fig.cap = "Depth profile (A) and bottom water warming of the Gulf of Maine. Everything deeper than 1000 meters is colored in green. B) Percentage change of bottom temperature relative to the present showing more intense warming in northern regions, especially under a high emission climate change scenario."}

@@

```{r ocFigureS24, eval = T, echo = F, fig.cap = "Projected MCP change relative to present (2005-2014) for each IPHC regulatory area. Colors represents the different ESM used in the study. Solid line represents a high emission scenario (RCP 8.5) and dashed line represents a low emission scenario (RCP 2.6)"}

@@

```{r ocFigure_out_outliers, eval = F, echo = F}
